Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-06-25 table_of_variables_and_constants.tex
Many small changes to all files -- GB final edits

  \startdata
  \multicolumn{2}{l}{\underline{Fluid Quantities}} \\[5pt]
  $E$ & comoving total fluid energy density \\
  $e$ & comoving thermal energy density \\
  $T$ & baryon temperature \\
  $S$ & fluid entropy \\
  $\rho$ & comoving baryonic density \\
  $\rho_{\!X}$ & comoving mass density of component $X$ (gas, dark matter, stars, metals, total, etc.) \\
  $t$ & time \\
  $\vecv$ & peculiar velocity \\
  $\vecB$ & comoving magnetic field strength \\
  $p$ & thermal pressure \\
  $p^*$ & total (thermal + magnetic) isotropic pressure \\
  $\gamma$ & ideal gas ratio of specific heats \\
  $\phi$ & gravitational potential \\
  $\Lambda$ & radiative/chemical cooling rate \\
  $\Gamma$ & radiative heating rate \\
  $Q$ & artificial viscosity stress tensor (only for Zeus hydro) \\[3pt]

@@

  \multicolumn{2}{l}{\underline{Natural Constants and Values}} \\[5pt]
  $G$ & gravitational constant \\
  $H$ & Hubble constant \\
  $k_B$ & Boltzmann constant \\
  $M_\odot$ & mass of Sun \\
  $m_H$ & mass of hydrogen atom \\
  $m_e$ & mass of electron \\
  $c$ & speed of light \\
  $c_{\rm s}$ & sound speed \\
  $z$ & redshift \\
  \enddata

>>>>

  \startdata
  \multicolumn{2}{l}{\underline{Fluid Quantities}} \\[5pt]
  $E$ & comoving total fluid energy density \\
  $e$ & comoving thermal energy density \\
  $T$ & baryon temperature \\
  $S$ & fluid entropy \\
  $\rho$ & comoving baryonic density \\
  $\rho_{\!X}$ & comoving mass density of component $X$ (gas, dark matter, stars, metals, total, etc.) \\
  $t$ & time \\
  $\vecv$ & peculiar velocity \\
  $\vecB$ & comoving magnetic field strength \\
  $p$ & comoving thermal pressure \\
  $p^*$ & comoving total (thermal + magnetic) isotropic pressure \\
  $\gamma$ & ideal gas ratio of specific heats \\
  $\phi$ & gravitational potential \\
  $\Lambda$ & radiative/chemical cooling rate \\
  $\Gamma$ & radiative heating rate \\
  $Q$ & artificial viscosity (only for \zeus\ hydro) \\[3pt]

@@

  \multicolumn{2}{l}{\underline{Natural Constants and Values}} \\[5pt]
  $G$ & gravitational constant \\
  $H$ & Hubble constant \\
  $k_B$ & Boltzmann constant \\
  $M_\odot$ & mass of Sun \\
  $m_H$ & mass of hydrogen atom \\
  $m_e$ & mass of electron \\
  $c$ & speed of light \\
  $c_{\rm s}$ & sound speed \\
  $z$ & redshift
  \enddata

### 2013-06-25 tests-onezonecollapse.tex
Many small changes to all files -- GB final edits

\begin{equation}
\frac{d\rho}{dt} = \frac{\rho}{t_{\rm col}},
\end{equation}

where the collapse timescale, $t_{\rm col}$, is

@@

\begin{equation}
\frac{de}{dt} = -p \frac{d}{dt} \frac{1}{\rho} - \Lambda,
\end{equation}

where $\Lambda$ is the cooling rate in units of erg s$^{-1}$ g$^{-1}$
and energy, temperature, density, and pressure are related by the
ideal gas law, including effects from molecular hydrogen as appropriate.
Figure \ref{fig.onezone} shows an example of the one-zone collapse test
performed with an initial number density of 1 hydrogen atom per cm$^{-3}$ and
an initial temperature of 100 K using the 12 species chemistry network with H,
D, and He species and metal cooling rates calculated with the \texttt{Cloudy}
code.  The effects of metal cooling can be clearly seen; as the metallicity
increases from zero to $10^{-2}$~Z$_\odot$, the gas rapidly and significantly
deviates from the primordial result (black line).  Our primordial results
compare very well to those shown in \citet{2005ApJ...626..627O}; however, we
use a different metal cooling method, so the lines describing the evolution of
the metal-enriched gas are not directly comparable.

>>>>

\begin{equation}
\frac{d\rho}{dt} = \frac{\rho}{t_{\rm col}},
\end{equation}
where the collapse timescale, $t_{\rm col}$, is

@@

\begin{equation}
\frac{de}{dt} = -p \frac{d}{dt} \frac{1}{\rho} - \Lambda,
\end{equation}
where $\Lambda$ is the cooling rate in units of erg s$^{-1}$ g$^{-1}$
and energy, temperature, density, and pressure are related by the
ideal gas law, including effects from molecular hydrogen as appropriate.
Figure \ref{fig.onezone} shows an example of the one-zone collapse test
performed with an initial number density of 1 hydrogen atom per cm$^{-3}$ and
an initial temperature of 100 K using the 12 species chemistry network with H,
D, and He species and metal cooling rates calculated with the \texttt{Cloudy}
code.  The effects of metal cooling can be clearly seen; as the metallicity
increases from zero to $10^{-2}$~Z$_\odot$, the gas rapidly and significantly
deviates from the primordial result (black line).  Our primordial results
compare very well to those shown in \citet{2005ApJ...626..627O}; however, we
use a different metal cooling method, so the lines describing the evolution of
the metal-enriched gas are not directly comparable.

### 2013-06-25 tests-orszagtang.tex
Many small changes to all files -- GB final edits

\subsubsection{MHD: Orszag-Tang Vortex}
\label{sec.tests.mhd}
Figure~\ref{fig.orszag} shows the Orszag-Tang vortex problem \citep{Orszag79}.  The left panel shows
the result using \enzo's constrained transport MHD method, while the right panel shows the result using
\enzo's implementation of Dedner MHD.
The Orszag-Tang vortex test is a classig MHD test problem, and shows that significant small scale structure can be generated in MHD
from large scale initial perturbations.  It is often used to compare the
effective resolution of different MHD schemes.  The test begins with uniform
density, $\rho_0=25/(36\pi)$, and pressure, $P_0=5/(12\pi)$ (as with
other tests in this section, in the absence of gravity or chemistry/cooling we use
arbitrary units).  There is a
single rotational mode in the velocity, and two in the magnetic field:
${\bf v}_0 = (-\sin(2\pi y) \hat{x},\, \sin(2\pi x) \hat{y})$,
${\bf B}_0 = (-\sin(2\pi y) \hat{x},\, \sin(4\pi x) \hat{y})$.
The simulation is evolved to $t=0.48$ (in arbitrary units).  One can see that the structures
are accurately represented as compared to, for example,
\citet{Toth00}, and that the resolution of shocks is comparable in
both methods.

>>>>

\subsubsection{MHD: Orszag-Tang Vortex}
\label{sec.tests.mhd}
Figure~\ref{fig.orszag} shows the Orszag-Tang vortex problem \citep{Orszag79}.  The left panel shows
the result using \enzo's constrained transport MHD method, while the right panel shows the result using
\enzo's implementation of Dedner MHD.
The Orszag-Tang vortex test is a classig MHD test problem, and shows that significant small scale structure can be generated in MHD
from large scale initial perturbations.  It is often used to compare the
effective resolution of different MHD schemes.  The test begins with uniform
density, $\rho_0=25/(36\pi)$, and pressure, $P_0=5/(12\pi)$ (as with
other tests in this section, in the absence of gravity or chemistry/cooling we use
dimensionless units).  There is a
single rotational mode in the velocity, and two in the magnetic field:
${\bf v}_0 = (-\sin(2\pi y) \hat{x},\, \sin(2\pi x) \hat{y})$,
${\bf B}_0 = (-\sin(2\pi y) \hat{x},\, \sin(4\pi x) \hat{y})$.
The simulation is evolved to $t=0.48$.  One can see that the structures
are accurately represented as compared to, for example,
\citet{Toth00}, and that the resolution of shocks is comparable in
both methods.

### 2013-06-25 tests-vandv.tex
Many small changes to all files -- GB final edits

After the test problems are selected by the test runner script, they
are run in succession using either the \enzo\ executable contained
within that distribution or an external executable built from another \enzo\
version and specified by the user.  This allows for any version of the
code to be tested with an identical set of test problems.  After
running the test problem simulations, the test runner then performs a
series of basic analysis tasks using the \texttt{yt} analysis toolkit
\citep{2011ApJS..192....9T, 2011arXiv1112.4482T}.  The default
analysis performed on all test problems includes calculation of
various statistics (such as extrema, mean, and variance) on the fields
present in the output data.  Custom analysis provided by scripts that
accompany the test problem parameter files is run for special cases,
such as when an analytical solution exists that can be compared
against the simulation data.  After the analysis is performed, the
results are compared against a set of gold standard results that are
maintained on a website and downloaded on the fly by the test runner
script.  Alternately, results from any version of \enzo\ can be stored
locally and compared to any other version of the code.  In
Section~\ref{sec.tests.problems}, we describe some of the key test
problems that are used to verify proper behavior.  All of these test
problems, as well as the scripts to generate the figures from them,
are available at \url{http://bitbucket.org/bwoshea/enzo-method-paper},
the Bitbucket repository for the \enzo\ method paper.  The
\texttt{test\_problems} subdirectory in this repository contains all of
the files necessary to regenerate all of the figures in
Section~\ref{sec.tests.problems}.  We note for completeness that 
the figures in Section~\ref{sec.tests.problems} were generated using \texttt{yt} version 2.5.3 and the
\url{https://bitbucket.org/enzo/enzo-dev} repository of \enzo\ with
changeset \texttt{5d8c412}.

@@

\enzo\ has been involved in multiple cosmological code comparisons,
including the Santa Barbara Cluster code comparison project
\citep{SantaBarbara}, a large comparison of N-body simulations
\citep{2008CS&D....1a5003H}, as well as several direct comparisons
between \enzo\ and the GADGET SPH code in a variety of
set-ups, including N-body and adiabatic hydrodynamics
\citep{2005ApJS..160....1O,2005MNRAS.364..909V, 2011MNRAS.418..960V}
and simulations of the Lyman-alpha forest \citep{2007MNRAS.374..196R}.
Compared to the other codes involved in these projects, the \enzo\ code typically has a more difficult time resolving
small-scale self-gravitating structures (for an equivalent dark matter
particle count and nominally equivalent force resolution), but does
comparably well as a tree-based code for larger structure, and is
typically superior in terms of resolving fluid features due to its
higher-order (and artificial viscosity-free) Piecewise Parabolic
Method hydrodynamics solver.  When examining classical test cases such
as the Santa Barbara project \citep{SantaBarbara}, \enzo\ forms galaxy
clusters with very similar density, temperature, and entropy profiles
to other grid-based codes that use Godunov-type hydro methods, which
systematically differed from particle-based codes using SPH in this
code comparison.  Similarly, in galaxy cluster simulations that look
specifically at the properties of cosmological shocks
\citep[e.g.][]{2011MNRAS.418..960V}, \enzo\ produces results that are
similar to other high-order grid-based hydrodynamics codes, and a far
superior performance of \enzo\ is observed (in terms of resolution of
fluid features and shock detection) in low-density regions when
compared to a particle-based code.  In tests of the Lyman-alpha forest
that include radiative cooling and a uniform metagalactic ultraviolet
background, \enzo\ and GADGET provide results on metrics such as the
matter power spectrum that are comparable to within 5\%
\citep{2007MNRAS.374..196R}.

Several comparisons have been made that focus specifically on
hydrodynamics solvers and fluid behavior.  In particular, the work of
\citet{2007MNRAS.380..963A} and \citet{Tasker2008} perform
direct comparisons between several grid- and particle-based
codes for a variety of fluid-centric test problems (including shocked gas clouds,
self-gravitating, translating clouds, and Sedov-Taylor explosions),
and show that \enzo\ is comparable or superior in behavior to the
other grid-based hydrodynamics codes involved in the comparison, and
provide useful information on the sort of practical challenges that a
user of an
AMR code such as \enzo\ may experience.  More specific comparisons, including one
specifically testing the linear and nonlinear growth of the
Kelvin-Helmholz instability \citep{2012ApJS..201...18M}, as well as a
comparison that more
broadly examines Galilean invariance in grid-based codes
\citep{2010MNRAS.401.2463R}, show that \enzo, and in particular its
implementation of the Piecewise Parabolic Method hydrodynamics solver,
converge to the correct solution as expected, and generally provide
less diffusive solutions than lower-order codes, including those that
use artificial viscosity.  Finally, there have been two code comparison
projects that focus on turbulence simulations.  The first studied the
behavior of decaying isothermal supersonic turbulence
\citep{2009A&A...508..541K}, and the second examined supersonic
magnetohydrodynamical turbulence \citep{2011ApJ...737...13K}.  Both
included the \enzo\ code, with the former testing the PPM
hydrodynamics and the latter both the constrained transport MHD
implementation of \citet{Collins10} and the Dedner method of \citet{WangAbelZhang08}.  In both cases, \enzo\ performed similarly to other
grid-based codes that use Godunov-based fluid solvers, and typically
had better effective resolution than particle-based codes when using the
same number of particles as the number of grid cells in the \enzo\ simulation.

Three other comparisons between the \enzo\ code and other simulation
tools have been performed that focus on physics other than gravity and
fluid flow.  The flux-limited diffusion radiation transport scheme was
measured against several test problems by \citet{IlievEtAl2009},
which involved tests with and without analytical solutions.  \enzo\ produced
results similar to both the analytical solutions and results obtained by 
other codes.  We note, however, that there were minor differences throughout
the comparison between various codes, and the majority of the codes differed
in at least a subset of the tests.
\citet{2011ApJ...726...55T} shows the results of varied reaction rate
coefficients for the formation of molecular hydrogen via the 3-body
process in both the \enzo\ and GADGET-2 codes, observing similar trends between
the two codes. However, at nominally equivalent resolution
where the particle and cell gas masses are comparable, 
\enzo\ simulations typically displayed a
substantially higher level of gas structure.  This is unsurprising due
to \enzo's higher-order hydro solver.  Finally,
\citet{2012ApJ...744...52P} show the results of comparing \enzo\ in
its non-AMR mode to a smoothed-particle hydrodynamics code, SNSPH, in
the context of common-envelope binary stellar interactions.  The
authors show that the codes display reasonable convergence properties
as a function of simulation resolution, and also agree quite well with
each other. However, the observed mass-loss rates do not agree
particularly well with observations.

>>>>

After the test problems are selected by the test runner script, they
are run in succession using either the \enzo\ executable contained
within that distribution or an external executable built from another \enzo\
version and specified by the user.  This allows for any version of the
code to be tested with an identical set of test problems.  After
running the test problem simulations, the test runner then performs a
series of basic analysis tasks using the \texttt{yt} analysis toolkit
\citep{2011ApJS..192....9T, 2011arXiv1112.4482T}.  The default
analysis performed on all test problems includes calculation of
various statistics (such as extrema, mean, and variance) on the fields
present in the output data.  Custom analysis provided by scripts that
accompany the test problem parameter files is run for special cases,
such as when an analytical solution exists that can be compared
against the simulation data.  After the analysis is performed, the
results are compared against a set of gold standard results that are
maintained on a website and downloaded on the fly by the test runner
script.  Alternately, results from any version of \enzo\ can be stored
locally and compared to any other version of the code.  In
Section~\ref{sec.tests.problems}, we describe some of the key test
problems that are used to verify proper behavior.  All of these test
problems, as well as the scripts to generate the figures from them,
are available at \url{http://bitbucket.org/enzo/enzo-method-paper},
the Bitbucket repository for the \enzo\ method paper.  The
\texttt{test\_problems} subdirectory in this repository contains all of
the files necessary to regenerate all of the figures in
Section~\ref{sec.tests.problems}.  We note for completeness that 
the figures in Section~\ref{sec.tests.problems} were generated using \texttt{yt} version 2.5.3 and the
\url{https://bitbucket.org/enzo/enzo-dev} repository of \enzo\ with
changeset \texttt{5d8c412}, which corresponds to \enzo\ 2.3.

@@

\enzo\ has been involved in multiple cosmological code comparisons,
including the Santa Barbara Cluster code comparison project
\citep{SantaBarbara}, a large comparison of N-body simulations
\citep{2008CS&D....1a5003H}, as well as several direct comparisons
between \enzo\ and the GADGET SPH code in a variety of
set-ups, including N-body and adiabatic hydrodynamics
\citep{2005ApJS..160....1O,2005MNRAS.364..909V, 2011MNRAS.418..960V}
and simulations of the Lyman-alpha forest \citep{2007MNRAS.374..196R}.
Compared to the other codes involved in these projects, the \enzo\ code typically has a more difficult time resolving
small-scale self-gravitating structures (for an equivalent dark matter
particle count and nominally equivalent force resolution), but does
comparably well as a tree-based code for larger structure, and is
typically superior in terms of resolving fluid features due to its
higher-order (and artificial viscosity-free) PPM hydrodynamics 
solver.  When examining classical test cases such
as the Santa Barbara project \citep{SantaBarbara}, \enzo\ forms galaxy
clusters with very similar density, temperature, and entropy profiles
to other grid-based codes that use Godunov-type hydro methods, which
systematically differed from particle-based codes using SPH in this
code comparison.  Similarly, in galaxy cluster simulations that look
specifically at the properties of cosmological shocks
\citep[e.g.][]{2011MNRAS.418..960V}, \enzo\ produces results that are
similar to other high-order grid-based hydrodynamics codes, and a far
superior performance of \enzo\ is observed (in terms of resolution of
fluid features and shock detection) in low-density regions when
compared to a particle-based code.  In tests of the Lyman-alpha forest
that include radiative cooling and a uniform metagalactic ultraviolet
background, \enzo\ and GADGET provide results on metrics such as the
matter power spectrum that are comparable to within 5\%
\citep{2007MNRAS.374..196R}.

Several comparisons have been made that focus specifically on
hydrodynamics solvers and fluid behavior.  In particular, the work of
\citet{2007MNRAS.380..963A} and \citet{Tasker2008} perform
direct comparisons between several grid- and particle-based
codes for a variety of fluid-centric test problems (including shocked gas clouds,
self-gravitating, translating clouds, and Sedov-Taylor explosions),
and show that \enzo\ is comparable or superior in behavior to the
other grid-based hydrodynamics codes involved in the comparison, and
provide useful information on the sort of practical challenges that a
user of an
AMR code such as \enzo\ may experience.  More specific comparisons, including one
testing the linear and nonlinear growth of the
Kelvin-Helmholz instability \citep{2012ApJS..201...18M}, as well as a
comparison that more
broadly examines Galilean invariance in grid-based codes
\citep{2010MNRAS.401.2463R}, show that \enzo, and in particular its
implementation of the PPM hydrodynamics solver,
converge to the correct solution as expected, and generally provide
less diffusive solutions than lower-order codes, including those that
use artificial viscosity.  Finally, there have been two code comparison
projects that focus on turbulence simulations.  The first studied the
behavior of decaying isothermal supersonic turbulence
\citep{2009A&A...508..541K}, and the second examined supersonic
magnetohydrodynamical turbulence \citep{2011ApJ...737...13K}.  Both
included the \enzo\ code, with the former testing the PPM
hydrodynamics and the latter both the constrained transport MHD
implementation of \citet{Collins10} and the Dedner method of \citet{WangAbelZhang08}.  In both cases, \enzo\ performed similarly to other
grid-based codes that use Godunov-based fluid solvers, and typically
had better effective resolution than particle-based codes when using the
same number of particles as the number of grid cells in the \enzo\ simulation.

Three other comparisons between the \enzo\ code and other simulation
tools have been performed that focus on physics other than gravity and
fluid flow.  The flux-limited diffusion radiation transport scheme was
measured against several test problems by \citet{IlievEtAl2009},
which involved tests with and without analytical solutions.  \enzo\ produced
results similar to both the analytical solutions and results obtained by 
other codes.  We note, however, that there were minor differences throughout
the comparison between various codes, and the majority of the codes differed
in at least a subset of the tests.
\citet{2011ApJ...726...55T} show the result of varying reaction rate
coefficients for the formation of molecular hydrogen via the 3-body
process in both the \enzo\ and GADGET-2 codes, observing similar trends between
the two codes. However, at nominally equivalent resolution
where the particle and cell gas masses are comparable, 
\enzo\ simulations typically displayed a
substantially higher level of gas structure.  This is unsurprising due
to \enzo's higher-order hydro solver.  Finally,
\citet{2012ApJ...744...52P} show the results of comparing \enzo\ in
its non-AMR mode to a smoothed-particle hydrodynamics code, SNSPH, in
the context of common-envelope binary stellar interactions.  The
authors show that the codes display reasonable convergence properties
as a function of simulation resolution, and also agree quite well with
each other. However, the observed mass-loss rates do not agree
particularly well with observations.

### 2013-06-25 tests-wavepool.tex
Many small changes to all files -- GB final edits

For the \zeus\ solver, we see that even before it has entered the refined region there are small oscillations excited behind the wave, although it is also worth noting that the wave itself is beautifully propagated without significant damping or phase errors.  In this test we use only our standard, low amount of quadratic artificial viscosity.  These waves could be damped by additional viscosity, but we do not add any in order to be sensitive to any artifacts at the grid boundary.  The remaining panels show that the trailing oscillations continue, but do not generate any additional noise. We note that the end result is similar to the case without any refined region.

Finally, the MUSCL solver shows also shows a very clean entry and exit from the refined region without any oscillations, although with mild damping on exit. However, because we use a piecewise linear reconstruction, the wave is spread more than with the other methods.

>>>>

For the \zeus\ solver, we see that even before it has entered the refined region there are small oscillations excited behind the wave, although it is also worth noting that the wave itself is beautifully propagated without significant damping or phase errors.  In this test we use only our standard, low amount of quadratic artificial viscosity.  The trailing oscillations could be damped by additional viscosity, but we do not add any in order to be sensitive to any artifacts at the grid boundary.  The remaining panels show that the trailing oscillations continue, but do not generate any additional noise. We note that the end result is similar to the case without any refined region.

Finally, the MUSCL solver also shows a very clean entry and exit from the refined region without any oscillations, although with mild damping on exit. However, because we use a piecewise linear reconstruction, the wave is spread more than with the other methods.
